---
format: html
editor: visual
---

-   **Section & Topic:**\[cite: 1\]
-   **Project Type:** (Implementation / Shiny App / Exploration / Simulation)\[cite: 1\]
-   **Question / Aim:**\[cite: 1\]
-   **Data Used:** (Include source details if real-world data)\[cite: 1\]
-   **Methods Used:**\[cite: 1\]

## Key Findings

<!-- Summary of findings -->

\[cite: 1\]

## Reflections

-   **What I would do differently next time:**\[cite: 1\]

## Links

-   [Detailed Analysis & Code Script](../appendix/code-and-data.qmd)\[cite: 1\]

```{r}
#|echo: false
#|warning: false
# Load required libraries: 
library(readr)
library(tidyverse)
library(tsibble)
library(lubridate)
library(dplyr)
```

```{r}
# Read in dataset: 
dat = read_csv('_data/ESK6816.csv')
# View(dat)
```

```{r}
# Convert data to time series object:
demand_ts = ts(
  dat$`RSA Contracted Demand`,
  start = c(2019, 2161), # Hour 2161 of 2019 (April 1st)
  frequency = 8766       # Hourly frequency per year
)

# Convert time series object to dataframe (for plotting):
demand_df = data.frame(
  hour    = as.numeric(time(demand_ts)),
  count   = as.numeric(demand_ts)
)

# Plot:
ggplot(data = demand_df,
       mapping = aes(x = hour,
                     y = count)) +  
  geom_line(col = 'deeppink')  +
  theme_minimal(base_size=14) +
  labs(title = 'South Africa hourly electricity demand.', 
       x = 'Time (hours)',
       y = 'Hourly peak demand')
```

As seen in the above plot, we cannot get a clear idea of the seasonality. We need to aggregate the data.

To handle the leap years present in the data we will use tsibble.

```{r}
# Create a tsibble dataframe:
df_ts = tibble(datetime=seq(from=as.POSIXct('2019-04-01 00:00:00'),
                              length.out=length(demand_ts),
                              by='hour'),
               demand=as.numeric(demand_ts)
               ) |>
  as_tsibble(index=datetime)

# Aggregate to daily values:
demand_daily_ts = df_ts |>
  index_by(date=~as.Date(.))|>
  summarise(demand_mean=mean(demand,na.rm=TRUE))

# Plot:
ggplot(data = demand_daily_ts,
       mapping = aes(x = date,
                     y = demand_mean)) +  
  geom_line(col = 'deeppink')  +
  theme_minimal(base_size=14) +
  labs(title = 'South Africa daily electricity demand.', 
       x = 'Time (days)',
       y = 'Daily peak demand')
```

Again, there is still a lot of unnecessary noise in the data so we're going to aggregate to months. 

```{r}
# Aggregate to monthly values:
demand_monthly_ts = df_ts |>
  index_by(year_month=~yearmonth(.))|>
  summarise(demand_mean=mean(demand,na.rm=TRUE))

# Plot:
ggplot(data = demand_monthly_ts,
       mapping = aes(x = year_month,
                     y = demand_mean)) +  
  geom_line(col = 'deeppink')  +
  theme_minimal(base_size=14) +
  labs(title = 'South Africa monthly electricity demand.', 
       x = 'Time (months)',
       y = 'Monthly peak demand')

```
